---
title: "Daily England Traffic"
description: "Analysis of average highway vehicle speed throughout the day in England from TidyTuesday"
author: Xander Nonaka
date: September 24, 2025 
format: html
execute:
  echo: false
  warning: false
  message: false
---

England National Highways Traffic Data Set

The data set used in this plot contains speed and vehicle volume data captured throughout the day over the course of many days on different highway sites in England.

```{r}
library(tidyverse)
library(dplyr)
```

```{r}
tuesdata <- tidytuesdayR::tt_load(2024, week = 49)

A64_traffic <- tuesdata$A64_traffic
```

```{r}
traffic <-A64_traffic %>% 
  group_by(`Time Interval`) %>% 
  summarize(avg_speed = mean(`Avg mph`, na.rm=TRUE)) %>% 
  arrange(desc(avg_speed))

ggplot(traffic, 
       aes(
         x=`Time Interval`,
         y=avg_speed,
       )) +
  geom_line() +
  labs(x="Time of Day",
       y="Average Speed (mph)",
       title="Highway Traffic Speed in England throughout the day")
```

Average speed across all highway sites is measured relative to the time of day the data was captured. Time is displayed here as a series of interval numbers from 0 to 95. 0 represents the first period of data on a new day, and 95 represents the last. Thus noon is around the 47th interval.

Sources:

National Highways, "WebTRIS Traffic Flow API." WebTRIS. https://webtris.nationalhighways.co.uk/api/swagger/ui/index

https://github.com/rfordatascience/tidytuesday/blob/main/data/2024/2024-12-03/readme.md
